---
title: "Impact of biostimulants on soil nitrogen dynamics"
author: "Mark Farrell"
date: "12/10/2018"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

# Overview

These data are from a pot experiment that was conducted to investigate whether biostimulants impact on soil nitrogen (N) cycling and wheat uptake of legume-derived N. Here, we are looking at soluble N and carbon (C) pools over the growth period. These are presented in mg L^-1^ as they are derived from soil solution samples that were collected non-destructively using [Rhizon samplers](https://www.rhizosphere.com/).

##### Code chunk to get web-based images into R markdown pdf output
```{r, fig.align='center'}
rhizon_url = 'https://www.rhizosphere.com/assets/images/RhizonCSSinacoreRRP.jpg'
if (!file.exists(rhizon <- 'RhizonCSSinacoreRRP.jpg'))
  download.file(rhizon_url, rhizon, mode = 'wb')
knitr::include_graphics(if (identical(knitr:::pandoc_to(), 'html')) rhizon_url else rhizon)
```

---

# Initial data processing, tidying, etc

First, we need to load tidyverse and the analysis packages, and bring the data into the R environment
```{r load_libraries, message=FALSE}
library(tidyverse)
library(lmerTest)
library(emmeans)
setwd("~/DATASCHOOL/r-learning/15N-continuation/")
raw_data <- read_csv("data/working/15n_soil_sol.csv")

```

We can now see that the data has been successfully loaded into the environment:

```{r raw}
raw_data
```

There are now several steps that need to be taken before we can start graphing the data. Some variables need to be calculated:

* DIN
* DON
* DOC_DON ratio
* DON_DIN ratio

## Making new variables

Making a new variable is easy. As I forgot to give it a name, I also re-named is using the ```name``` function in base R

```{r DIN, message=FALSE}
with_din <- mutate(raw_data, nh4 + no3)
names(with_din)[11] <- "din"
```

## Data cleaning
As part of this there is some cleaning that also needs to happen. In calculating DON, we can use ```case_when()``` in the mutate function to identify values < 0 and make them 0:

```{r mutations, message=FALSE}
with_don <- with_din %>% 
  mutate(don = tdn - din,
         don_nz = case_when(
           don >= 0 ~ don,
           TRUE ~ 0
         ))
```

```{r hidden 1, echo=FALSE, message=FALSE, results=FALSE}
##do we have any negatives?
count(with_don, don_nz < 0)

##tidying...
don_fixed <- select(with_don, -don)
names(don_fixed)[12] <- "don"
```

Doing some more simple mutations calculates the ratios, but also generates ```Inf``` where a divide by zero has happened

```{r ratios_calc}
ratios <- mutate(don_fixed, doc_don = doc / don )
ratios <- mutate(ratios, don_din = don / din )
```

```{r ratios, echo=FALSE}
ratios
```

The ```case_when``` operation above won't work here straight away to turn ```Inf``` to ```'NA'```

So, Alex provided a solution using <mutate> to change all infinate values to NA:
```don_fixed$doc_don[which(is.infinite(don_fixed$doc_don))] <- NA
```
I also found one on [Stack Overflow](https://stackoverflow.com/questions/12188509/cleaning-inf-values-from-an-r-dataframe)

```{r kill_inf, message=FALSE}
is.na(ratios) <- do.call(cbind,lapply(ratios, is.infinite))
```

This searches the data frame, finds infinate values and replaces them with NA

```{r, echo=FALSE}
ratios
```

Finally, the finished dataframe is written to a new *.csv

```write_csv(ratios, "data/processed/ratios.csv")```

# Data exploration with `ggplot`

```{r hidden , echo=FALSE, message=FALSE, results=FALSE}
setwd("~/DATASCHOOL/r-learning/15N-continuation/")
raw_data <- read_csv("data/working/ratios.csv")
```

Given we have time-series data, it seems most appropriate to explore the different variables with time in days as the x-axis.

## Graphing concepts

* You can include several objects with the aesthetic `aes()`. The format is x-axis, y-axis, ...
    * Where ... can include colour, shape, size, etc
* You can set colurs manually using `scale_colour_manual(values=c("#CCFF00"))` where colours are provided in hexadecimal code in the order in which the variable is set
* `geom_point` or `geom_jitter` can be used for dot plots. The jitter option allows the points to be moved slightly in the x-axis to reduce over-plotting, which is what we've chosen here given the large number of data points
* `stat_smooth` fits lines through the points. We've used loess here as an accepted way of interpolating between the points
* `coords_cartesian` allows soft x and y limits
* Given the awkwardness in coding special symbols, sub/super-script, brackets etc in axis lables, I found a solution using `bquote`. So, for more polished plots, rather than defining x and y within the `labs()` function, these are defined first as `xlab(bquote())`. `labs()` can then just be used for the title, legends, etc.
* Theme is then used to tidy things up

There's definitely lots more to this, and teh R graphics cookbook is a good start for different graph types. Needless to say, you can spend rather too long on your artwork at this stage, so something to keep an eye on!

## What are we looking for in the data?

In a nutshell, we're particularly interested in how the variables measured co-vary over time as affected by treatment. But, as this has the potential to get very messy, we'll first produce some simpler figures that will help us pin-point some main themes.

### Variation with time

1. Nitrate concentrations:

```{R, echo=FALSE, fig.align='center', out.width = "75%"}
ggplot(data = ratios, aes(days, no3, colour = trt)) + # x, y, colour by treatment
  scale_colour_manual(values=c("#CCFF00", # all the colours in hexadecimal code in order of "colour = x" argument above
                               "#99FF00",
                               "#66FF00",
                               "#00CC00",
                               "#336600",
                               "#FFCC33",
                               "#FF9933",
                               "#CC9933",
                               "#996600",
                               "#993300",
                               "#333333",
                               "#0099FF"
  )) +
  geom_jitter(size = 4, alpha = 0.7, width = 0.5) + # Plot as a jittered point graph
  stat_smooth(method = "loess", se = FALSE, size = 1) + # Adds trend line only
  coord_cartesian(xlim = c(0, 48), ylim = c(0, 450)) + #sets soft limits to scale
  labs(
    x = "Time (days)", # x axis label
    y = ~NO[3]^{textstyle("-")}-N ~mg~L^{-1}, #y axis label, note cpde and weblink above for special characters and script
    colour = "Treatment" # Legend Title
  ) +
  theme(
    panel.grid = element_blank(), # blank grid
    axis.text=element_text(size=14), # axis text size
    axis.title = element_text(color="black", face="bold", size=18), # axis label text size
    panel.background = element_rect(fill = "white", # panel background
                                    colour = "black") # panel outline
  )

```

2. Ammonium concentrations:

```{R, echo=FALSE, fig.align='center', out.width = "75%"}
ggplot(data = ratios, aes(days, nh4, colour = trt)) + # x, y, colour by treatment
  scale_colour_manual(values=c("#CCFF00", # all the colours in hexadecimal code in order of "colour = x" argument above
                               "#99FF00",
                               "#66FF00",
                               "#00CC00",
                               "#336600",
                               "#FFCC33",
                               "#FF9933",
                               "#CC9933",
                               "#996600",
                               "#993300",
                               "#333333",
                               "#0099FF"
  )) +
  geom_jitter(size = 4, alpha = 0.7, width = 0.5) + # Plot as a jittered point graph
  stat_smooth(method = "loess", se = FALSE, size = 1) + # Adds trend line only
  labs(
    x = "Time (days)", # x axis label
    y = ~NH[4]^{textstyle("+")}-N ~mg~L^{-1}, #y axis label, note cpde and weblink above for special characters and script
    colour = "Treatment" # Legend Title
  ) +
  theme(
    panel.grid = element_blank(), # blank grid
    axis.text=element_text(size=14), # axis text size
    axis.title = element_text(color="black", face="bold", size=18), # axis label text size
    panel.background = element_rect(fill = "white", # panel background
                                    colour = "black") # panel outline
  )

```

3. Dissolved organic nitrogen concentrations

```{R, echo=FALSE, fig.align='center', out.width = "75%"}
ggplot(data = ratios, aes(days, don, colour = trt)) + # x, y, colour by treatment
  scale_colour_manual(values=c("#CCFF00", # all the colours in hexadecimal code in order of "colour = x" argument above
                               "#99FF00",
                               "#66FF00",
                               "#00CC00",
                               "#336600",
                               "#FFCC33",
                               "#FF9933",
                               "#CC9933",
                               "#996600",
                               "#993300",
                               "#333333",
                               "#0099FF"
  )) +
  geom_jitter(size = 4, alpha = 0.7, width = 0.5) + # Plot as a jittered point graph
  stat_smooth(method = "loess", se = FALSE, size = 1) + # Adds trend line only
  labs(
    x = "Time (days)", # x axis label
    y = DON ~mg~L^{-1}, #y axis label, note cpde and weblink above for special characters and script
    colour = "Treatment" # Legend Title
  ) +
  theme(
    panel.grid = element_blank(), # blank grid
    axis.text=element_text(size=14), # axis text size
    axis.title = element_text(color="black", face="bold", size=18), # axis label text size
    panel.background = element_rect(fill = "white", # panel background
                                    colour = "black") # panel outline
  )

```

4. Dissolved organic carbon - a potentially useful co-variate

```{R, echo=FALSE, fig.align='center', out.width = "75%"}
ggplot(data = ratios, aes(days, doc, colour = trt)) + # x, y, colour by treatment
  scale_colour_manual(values=c("#CCFF00", # all the colours in hexadecimal code in order of "colour = x" argument above
                               "#99FF00",
                               "#66FF00",
                               "#00CC00",
                               "#336600",
                               "#FFCC33",
                               "#FF9933",
                               "#CC9933",
                               "#996600",
                               "#993300",
                               "#333333",
                               "#0099FF"
  )) +
  geom_jitter(size = 4, alpha = 0.7, width = 0.5) + # Plot as a jittered point graph
  stat_smooth(method = "loess", se = FALSE, size = 1) + # Adds trend line only
  labs(
    x = "Time (days)", # x axis label
    y = DOC ~mg~L^{-1}, #y axis label, note cpde and weblink above for special characters and script
    colour = "Treatment" # Legend Title
  ) +
  theme(
    panel.grid = element_blank(), # blank grid
    axis.text=element_text(size=14), # axis text size
    axis.title = element_text(color="black", face="bold", size=18), # axis label text size
    panel.background = element_rect(fill = "white", # panel background
                                    colour = "black") # panel outline
  )

```

These four figures show some interesting trends. Nitrate peaks at 14 days, whereas ca. 100-fold lower ammonium concentrations are at their highest upon commencement of the experiment, and are much lower for the zero control (treatment L). Dissolved organic nitrogen data is generally noiser, but the solid treatments (F-J, browns) seem to result in evelated DON concentrations again around d14. DOC becomes most elevated by treatment B at d48 (harvest).

### Integration...

Given the dominance of nitrate within the dissolved nitrogen pool, it perhaps makes sense to have this as a focus. However, as it's not the only form of avilable N measured, I have chosen to plot total dissolved N, with spot size representing the proportion of that N which is nitrate. This has required a couple more data processing steps:

```{R, echo=TRUE}
no3percent <- mutate(ratios, percentno3 = (100 / tdn) * no3 )
no3fix <- mutate(no3percent, if_else(percentno3 >100, 100, percentno3))
no3fix <- rename(no3fix, percentno3f = "if_else(percentno3 > 100, 100, percentno3)")
no3fix
```

This has made a new variable `percentno3f` within a new tibble `no3fix`. It is from this that we can now construct the final time-based figure, before moving on to analysis of teh data at key time points. This time I've echoed the `ggplot` code so that we can see how it's constructed:

```{R, echo=TRUE, fig.align='center',out.width = "90%"}
ggplot(data = no3fix, aes(days, tdn, colour = trt, size = percentno3f)) + # x, y, colour by treatment
  scale_colour_manual(values=c("#CCFF00", # all the colours in hexadecimal code in order of "colour = x" argument above
                               "#99FF00",
                               "#66FF00",
                               "#00CC00",
                               "#336600",
                               "#FFCC33",
                               "#FF9933",
                               "#CC9933",
                               "#996600",
                               "#993300",
                               "#333333",
                               "#0099FF"
  )) +
  geom_jitter(alpha = 0.7, width = 0.5) + # Plot as a jittered point graph
  stat_smooth(method = "loess", se = FALSE, size = 1) + # Adds trend line only
  coord_cartesian(xlim = c(0, 48), ylim = c(0, 520)) + #sets soft limits to scale
  xlab(bquote('Time ('*d*')')) +
  ylab(bquote('TDN ('*'mg N'%.% L^-1*')')) +
  labs(
    title = "Total dissolved N in soil solution", # Graph title
    colour = "Treatment", # Legend Title
    size = "%" ~NO[3]^{textstyle("-")}
  ) +
  theme(
    plot.title = element_text(size=22),
    panel.grid = element_blank(), # blank grid
    axis.text=element_text(size=14), # axis text size
    axis.title = element_text(color="black", face="bold", size=18), # axis label text size
    panel.background = element_rect(fill = "white", # panel background
                                    colour = "black") # panel outline
  )

```

This figure shows us a lot of useful information:

* Dissolved N at the start of the pot experiment is typically dominated by non-nitrate forms
* By the peak at day 14, the majority of dissolved N in all treatments takes the form of nitrate
* This falls rapidly by day 36
* There is an interesting up-tick in TDN in one of the liquid treatments by day 48 (harvest) that is not related to nitrate. Looking at the initial exploratory figures, it is apparent that this is dissolved organic N, reflective of an increase in dissolved organic C. The mechanisms for this are unknown, but appear to only be related to this one product

---



Then split data

Then LMERTEST etc

Check alll works neatly, progress to plant data if time
