---
title: "Chapter 1"
author: "Mark Farrell"
date: "09/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("emmeans")
```

Chapter 1 starts on t-tests then simple linear models for the wheat dataset from agridat. 

First task is to sort data and get it ready for analysis:

```{r data_sort, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry")
wheat <- read_csv("data/working/wheat yield.csv")
as_factor(wheat$Variety)
wheat$Variety <- factor(wheat$Variety, levels = c("Standard", "New"))
wheat_H <- filter(wheat, Variation == "High")
wheat_L <- filter(wheat, Variation == "Low")

```

A simple plot lets us see the data


```{r out.width = "50%", echo=TRUE}
ggplot(wheat, aes(Variety, Yield, colour=Variety)) + 
  geom_point() + 
  facet_wrap(~Variation)
```


Now data is all sorted and we've had a peek at what it looks like, simple t-tests run easily. 

```{r t-testing, echo=TRUE}
t.test(Yield~Variety, data = wheat_H, var.equal=TRUE)
t.test(Yield~Variety, data = wheat_L, var.equal=TRUE)
```

As can be seen, there's no significant effect of Variety in the high variance half of the dataset, but in the low variance half, _P_ << 0.05

Given that t-tests are a bit pasé, and also ultimately inappropriate for a dataset like this (noting trhere are really two factors), we'll explore linear modelling.

```{r, echo=TRUE}
lm1 <- lm(Yield ~ Variety, data = wheat_L)
anova(lm1)
```

Here we see that for the low variance half of the dataset, there is a significant effect of variety. To look closer at the fitted model, we'll run `summary(lm1)` to give info on the residuals and coefficients, goodness of fit, etc. and `emmeans(lm1, ~Variety)` to provide output of estimated marginal means of model.
```{r}
summary(lm1)
emmeans(lm1, ~Variety)
```
The final task is to plot the output of the model. This took a bit of wrangling, but a passable ggplot output solution is presented below:

```{r echo=TRUE, out.width = "50%"}
lm1.results<-summary(emmeans(lm1,~Variety))

ggplot(lm1.results, aes(Variety, emmean, fill=Variety)) +
  geom_bar(stat="identity", width=.4) +
  geom_errorbar(aes(ymin = emmean-SE, ymax = emmean+SE), width=.2) + 
                  ylim(0,5) +
                  labs(y = 'Mean Yield (t/hectare)') 
```

Going though the structure of this ggplot command, a few things are happening:

1. First we make a new data frame from the summary outputs of `emmeans`
2. This is then plotted with Variety on the x-axis and emmean on the y-axis, which is the output of the model. Fill is purely for prettiness
3. `stat="identity"` tells R to use the y values for the height of the bars
4. Error bars are calculated from the emmean +/- SE within the `lm1.results` data frame
5. Y axis limits are imposed
6. The proper y-axis label is then added. Note this was very particular and didn't like double quotes. There's lots of info on getting sub/superscript, symbols, etc. on axis titles available. Best searched on a case-by-case basis.

