---
title: "Chapter 4"
author: "Mark Farrell"
date: "11/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("emmeans")
library("lmerTest")
```

# Part 1 - How to incorporate components of a statistical model into the model statement

* Blocking factors can be a significant source of variation
* Between-block variation can interfere with treatment comparison
* Treatments comparison ---- within blocks
* Separate out Between-block variation  within-block variation

## Example 1: Is photosynthetic rate affected by temperature?

```{r data_sort, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry/")
photosynthesis<-read_csv("data/working/Prac 4 photosynthesis.csv")
photosynthesis$Position<-factor(photosynthesis$Position)
photosynthesis$Temp<-factor(photosynthesis$Temp)
```

24 plants grown in 3 different parts of the greenhouse

Outcome measure
      Photosynthetic rate
Experimental factors
      Temperature (high/low)
Blocking factors
      Position (4)

_2 parameter model (plus random effects)_

```{R, out.width = "50%", echo=TRUE}
ggplot(photosynthesis,aes(Temp,PhotoRate,colour=Position))+
  geom_point()
```

To demonstrate the point, first looking at a "normal" model without the blocking factor, we see that temperature has no effect on photosynthetic rate, which doesn't really make a whole lot of sense.

```{R, echo=TRUE}
model1=lm(PhotoRate~Temp, data=photosynthesis)
anova(model1)
```

So, by using `lmer` from `lmerTest`, we're able to bring in teh random factor of block withthe following argument `+(1|Position)` where `Position` represents the block in this case.

```{R, echo=TRUE}
model2=lmer(PhotoRate~Temp+(1|Position), data=photosynthesis)
anova(model2)
```

We now have a significant impact of temperature on photosynthetic rate. exploring the contrasts between teh two models by examining the summaries,

```{R, echo=TRUE}
summary(model1)
summary(model2)
```

it's clear how much teh standartd error around the temperature effect tightens up once the random effect of position is accounted for.

```{R, echo=TRUE}
emmeans(model2,~Temp)
```

Finally, assumption checking...


```{R, out.width = "50%", echo=TRUE}
plot(model2)
```

## Example 2 - Can a gene KO Arabidopsis modulate leaf temperature during drought?

Outcome measure
        Leaf temperature
Experimental factors
       Genotype (control/mutant) (2)
        Watering conditions (2)
Blocking factors
        Plant (4)
        
_4 parameter model (plus random effects)_

```{r, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry/")
drought<-read_csv("data/working/Prac 4 drought data.csv")
str(drought)

drought$plant<-factor(drought$plant)

drought$Genotype <- factor(drought$Genotype, levels = c("WT", "mutant"))
drought$WaterCondition <- factor(drought$WaterCondition, levels = c("Normal", "Drought"))
```

The data is now imported, columns set as factors where appropriates, and the "reference" levels set with wild-type for Genotype, and Normal for WaterCondition.

Next, to examine the data, we'll put the interaction between genotype and water condition on the x-axis, then do a dot plot as usual.

```{R, out.width = "50%", echo=TRUE}
drought$cond<-with(drought,interaction(WaterCondition, Genotype))
ggplot(drought, aes(cond, Temperature,colour=plant))+geom_point()
```

As a demonstration that the proper inclusion of variance structure is vital to successful analysis, we'll do a standard linear model first:

```{R, echo=TRUE}
lm.drought<-lm(Temperature~Genotype*WaterCondition, data=drought)
anova(lm.drought)
```

This shows a strong impact of Genotype, Water Condition and the two-way interaction. but it might not be correct.... Let's see what happens with a more appropriate mixed model including Plant as a blocking random factor:

```{R, echo=TRUE}
lmer.drought<-lmer(Temperature~Genotype*WaterCondition+(1|plant), data=drought)
anova(lmer.drought)
```

So, what's happening? The clue is within `emmeans`:

```{R, echo=TRUE}
emmeans(lm.drought,~Genotype*WaterCondition)
emmeans(lmer.drought,~Genotype*WaterCondition)
```

As the linear model knows nothing about the blocking factor, it assigns all 28 degrees of freedom to each factor. The mixed model however recognises this as erroneous, and correctly allocates just 4 df to each factor, resulting in much larger standard error estimates, and thus far less clarity on effect.
