---
title: "Chapter 4"
author: "Mark Farrell"
date: "11/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("emmeans")
library("lmerTest")
```

# Part 1 - How to incorporate components of a statistical model into the model statement

* Blocking factors can be a significant source of variation
* Between-block variation can interfere with treatment comparison
* Treatments comparison ---- within blocks
* Separate out Between-block variation  within-block variation

## Example 1: Is photosynthetic rate affected by temperature?

```{r data_sort, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry/")
photosynthesis<-read_csv("data/working/Prac 4 photosynthesis.csv")
photosynthesis$Position<-factor(photosynthesis$Position)
photosynthesis$Temp<-factor(photosynthesis$Temp)
```

24 plants grown in 3 different parts of the greenhouse

Outcome measure
      Photosynthetic rate
Experimental factors
      Temperature (high/low)
Blocking factors
      Position (4)

_2 parameter model (plus random effects)_

```{R, out.width = "50%", echo=TRUE}
ggplot(photosynthesis,aes(Temp,PhotoRate,colour=Position))+
  geom_point()
```

To demonstrate the point, first looking at a "normal" model without the blocking factor, we see that temperature has no effect on photosynthetic rate, which doesn't really make a whole lot of sense.

```{R, echo=TRUE}
model1=lm(PhotoRate~Temp, data=photosynthesis)
anova(model1)
```

So, by using `lmer` from `lmerTest`, we're able to bring in teh random factor of block withthe following argument `+(1|Position)` where `Position` represents the block in this case.

```{R, echo=TRUE}
model2=lmer(PhotoRate~Temp+(1|Position), data=photosynthesis)
anova(model2)
```

We now have a significant impact of temperature on photosynthetic rate. exploring the contrasts between teh two models by examining the summaries,

```{R, echo=TRUE}
summary(model1)
summary(model2)
```

it's clear how much teh standartd error around the temperature effect tightens up once the random effect of position is accounted for.

```{R, echo=TRUE}
emmeans(model2,~Temp)
```

Finally, assumption checking...


```{R, out.width = "50%", echo=TRUE}
plot(model2)
```

## Example 2 - Can a gene KO Arabidopsis modulate leaf temperature during drought?

Outcome measure
        Leaf temperature
Experimental factors
       Genotype (control/mutant) (2)
        Watering conditions (2)
Blocking factors
        Plant (4)
        
_4 parameter model (plus random effects)_

```{r data_sort, echo=TRUE, results="hide"}
drought<-read_csv("data/working/Prac 4 drought data.csv")
str(drought)

drought$plant<-factor(drought$plant)

drought$Genotype <- factor(drought$Genotype, levels = c("WT", "mutant"))
drought$WaterCondition <- factor(drought$WaterCondition, levels = c("Normal", "Drought"))
```

The data is now imported, columns set as factors where appropriates, and the "reference" levels set with wild-type for Genotype, and Normal for WaterCondition.