---
title: "Chapter 5"
author: "Mark Farrell"
date: "11/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("emmeans")
library("lmerTest")
```

# Statistical modelling: Understanding Error (Variance) Structure II

## Part 1 - Getting the error term correct: a case study 

Main research questions: 

* Does dark respiration vary within a leaf?
* How does dark respiration vary by leaf age?  
* Does dark respiration vary across species? 

Measurements: 

* 9 plants- 3 plants/species (3 species)
* 3 leaf stages (ages): older - younger leaves
* 5 leaf sections from base to tip
* n = 9x3x5 = *135 observations*

```{r, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry/")
respire<-read_csv("data/working/Prac 4 dark respiration.csv")
respire$Plant_ID<-factor(respire$Plant_ID)
respire$Leaf_stage<-factor(respire$Leaf_stage)
respire$Leaf_section<-factor(respire$Leaf_section)
respire$Species<-factor(respire$Species, labels=c("PM","UP","ZM"))
```

All data imported, appropriate columns set as factors, Species factor re-ordered / ne-named. Now lets have a quick look at the data:

```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
ggplot(respire, aes(Species, Dry_mass_resp,colour=Leaf_section))+geom_boxplot()
ggplot(respire, aes(Species, Dry_mass_resp,colour=Leaf_stage))+geom_boxplot()
```

There's an important structure to the variance of this data.

1. We have the fully factorial `Species x Leaf Stage x Leaf section` portion of the model
2. We then have the plant as a block `+(1|Plant_ID)`
3. We then have leaf stage nested within plant block `+(1|Leaf_stage:Plant_ID)`

```{R, echo=TRUE}
model4<-lmer(Dry_mass_resp~Species*Leaf_stage*Leaf_section+(1|Plant_ID) + (1|Leaf_stage:Plant_ID),data=respire) 
anova(model4)
```

We can conlude that only leaf section has a significant impact on respiration.

```{R, out.width = "50%", echo=TRUE}
plot(model4)
```

Fit is good too :)