---
title: "Chapter 3"
author: "Mark Farrell"
date: "10/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("emmeans")
```

# Part 1 - Can drought tolerance in sorghum be improved through genetic modification?

This exercise explores two-factpor ANOVA. First, we need to import the data.

```{r data_sort, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry")
data2 <- read_csv("data/working/Prac 3 mock LWR.csv")
```

With data imported, a simple plot is produces to both show all the data (dots) and its variance structure (b&w):

```{R, echo=TRUE, out.width = "50%"}
ggplot(data2, aes(GeneB,LWR,colour=GeneA)) + 
  geom_boxplot() + 
  geom_point(position=position_dodge(0.75))
```

Now, we can fit a two-factor full factorial model to the data, and test for significance

```{R, echo=TRUE}
lm1<-lm(LWR~GeneA*GeneB, data = data2)
anova(lm1)
```

As can be seen, both genes AA and BB have a significant impact on leaf water retention (LWR), and there is also a slightly significan interaction effect. To probe this further, we can then output the model summary, and ask emmeans to produce pairwise comparison outputs.

```{R, echo=TRUE}
summary(lm1)
emmeans(lm1, pairwise~GeneA|GeneB)
emmeans(lm1, pairwise~GeneB|GeneA)
```

This outputs all the multiple comparisons, allowing us to see if bb == or =/= BB when aa or AA are present, and vice versa. 

Finally, we need to check the model assumptions using the simple outputs provided by `plot`.

```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
plot(lm1, which=1) 
plot(lm1, which=2)
```

All looks good :-)

# Part 2 Which cabbage cultivar has the higher Vitamin C content on average?

Here, we're exploring model selection. We have two cabbage cultivars (factor 1) planted on three different days (factor 2)

```{R, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry")
cabbage<-read_csv("data/working/Prac 3 cabbage data.csv")
```

Plot to see the data:

```{R, echo=TRUE, out.width = "50%"}
ggplot(cabbage,aes(Date,VitC,colour=Cult))+
  geom_boxplot() +
  geom_point(position=position_dodge(0.75))
```

First, we'll use a full factorial design to confirm no interaction between planting date and cultivar:

```{R, echo=TRUE}
lm2<-lm(VitC~Cult*Date, data = cabbage)
anova(lm2)
```

As it can be seen that there isn't an interaction, a more appropriate model is an additive model. This is a better test of the question: "Does cultivar affect vitamin C levels?"

```{R, echo=TRUE}
lm3<-lm(VitC~Cult+Date, data = cabbage)
anova(lm3)
emmeans(lm3, pairwise~Cult)
```

It's clear that c39 has lower vitamin C levels than c52, by 12.9 +/- 1.8 units.
```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
plot(lm3, which=1) 
plot(lm3, which=2)
```

Finally, assumptions for ANOVA look OK. Crack on...

# Part 3 - model choices...

Experimental set-up: 
* 60 plots
* 5 varieties: “M”, “P”, “S”, “T”, “V”
* 6 locations: “C”,”D”, “GR”, “M”, “UF”, “W”
* 2 crop years: 1930, 1931

Research question: Compare yields of 5 varieties of barley

```{R, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry")
barley<-read_csv("data/working/Prac 3 barley yield.csv")
```

As this data is untidy (Y1 and Y2 need to be in long-form), we need to use the `gather` function in the tidyverse (https://tidyr.tidyverse.org/reference/gather.html) to fix things up.

```{R, echo=TRUE}
barley2<-gather(barley, key="year", value="yield", 4:5)
str(barley2)
```

Here, we're making a new tibble `barley2` from `barley`. Once this has been achieved, we can plot the data a couple of different ways to look at whether or not year is important when assessing the impact of barley variety on yield.

```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
ggplot(barley2,aes(Var,yield,colour=year))+
  geom_boxplot() +
  geom_point(position=position_dodge(0.75))
ggplot(barley2,aes(year,yield,colour=Var))+
  geom_boxplot() +
  geom_point(position=position_dodge(0.75))
```

From these, it's pretty clear there isn't an interaction with year at the whole dataset level, with yields likely significantly higher in Y1 than Y2.

First we'll try with location and variation in an additive model, but won't feed the model any info on year:

```{R, echo=TRUE}
lm4<-lm(yield~Var+Loc, data = barley2)
anova(lm4)
```

This retuns a strongly significant impact of location on yield (not the question), but only slightly significant impact of variety (the actual question). Perhaps if we include year in the model this will help address some of teh unexplained variance and we might get a clearer picture of how important variety is.

```{R, echo=TRUE}
lm5<-lm(yield~Var+Loc+year, data = barley2)
anova(lm5)
```

Yes, as predicted, year was significant, and its inclusion has actually increased the significance of the variance result as this has dealt with some of the unexplained variance in `lm4`. Now let's see which varieties are actually different:

```{R, echo=TRUE}
emmeans(lm5, pairwise~Var)
```

Here, we can see that M and S had a significantly lower yield than T, but all other varieties had limited impact on yield

```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
plot(lm5, which=1) 
plot(lm5, which=2)
```

All looks good

# Part 4 - Relationships between two variables as affected by a factor
Research question: What is the relationship between mean quadratic diameter and plot density in Nothofagus forest in the Andes? Does this relationship differ between tree species?

Experimental set-up: 

* 41 plots
* 3 species (StandTypes)
* Outcome measure: Plot Density
* Factors: StandType, QuadDiam
* Blocking factors: ?

```{R, echo=TRUE, results="hide"}
setwd("~/DATASCHOOL/r-learning/stats-terry")
forest<-read_csv("data/working/Prac 3 forest.csv")
```

Quick visual exploration of the data reveals a general trend of decreased quadratic diameter with increased density. Is this affected by tree species?

```{R, echo=TRUE, out.width = "50%"}
ggplot(forest,aes(Density,QuadDiam,colour=StandType))+
  geom_point()
```

Is this affected by tree species?

```{R, echo=TRUE, out.width = "50%"}
ggplot(forest,aes(Density,QuadDiam,colour=StandType))+
  geom_point() +
  stat_smooth(method=lm, se=FALSE)
```

Possibly.... So, let's try a model with the interaction term to test that assertion:

```{R, echo=TRUE}
lm6<-lm(Density~QuadDiam*StandType, data = forest)
anova(lm6)
```

As we can see, whilst both Density and QuadDiam are strongly significant, there is no sign of a significant interaction, thus a more appropriate additive model can be used:

```{R, echo=TRUE}
lm7<-lm(Density~QuadDiam+StandType, data = forest)
anova(lm7)
```

Stand type is strongly significant, calling emmeans to conduct the Tukey's multiple comparisons...

```{R, echo=TRUE}
emmeans(lm7, pairwise~StandType)
```

We can see here Coigue is different to both Rauli and Roble, but Rauli and Roble do not differ from each other. Let's just confirm that the model assumptions aren't too awry, and we're done here.

```{R, echo=TRUE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.show='hold',fig.align='center'}
plot(lm7, which=1) 
plot(lm7, which=2)
```